\section*{Contributor's list}

The following are some of the people who have contributed to this
book:

\begin{itemize}

\item Many of the problems in this book are variations of classical
problems that appeared first in technical articles and then in textbooks.
Whenever I know the origin of a problem or solution, I acknowledge it
in the text.

\item I also thank the students at Wellesley College who worked with
the first edition of the book, and the students at Olin College who
worked with the second edition.

\item Se Won sent in a small but important correction in my presentation
of Tanenbaum's solution to the Dining Philosophers Problem.

\item Daniel Zingaro punched a hole in the Dancer's problem, which
provoked me to rewrite that section.  I can only hope that it makes more
sense now.  Daniel also pointed out an error in a previous version of
my solution to the H$_2$O problem, and then wrote back a year later
with some typos.

\item Thomas Hansen found a typo in the Cigarette smokers problem.

\item Pascal R\"{u}tten pointed out several typos, including my embarrassing
misspelling of Edsger Dijkstra.

\item Marcelo Johann pointed out an error in my solution to the
Dining Savages problem, and fixed it!

\item Roger Shipman sent a whole passel of corrections as well as
an interesting variation on the Barrier problem.

\item Jon Cass pointed out an omission in the discussion of dining
philosophers.

\item Krzysztof Ko\'{s}ciuszkiewicz sent in several corrections, including
a missing line in the Fifo class definition.

\item Fritz Vaandrager at the Radboud University Nijmegen in the
Netherlands and his students Marc Schoolderman, Manuel Stampe and Lars
Lockefeer used a tool called UPPAAL to check several of the solutions
in this book and found errors in my solutions to the Room Party problem
and the Modus Hall problem.

\item Eric Gorr pointed out an explanation in Chapter 3 that was
not exactly right.

\item Jouni Lepp\"{a}j\"{a}rvi helped clarify the origins of semaphores.

\item Christoph Bartoschek found an error in a solution to
the exclusive dance problem.

\item Eus found a typo in Chapter 3.

\item Tak-Shing Chan found an out-of-bounds error in {\tt counter\_mutex.c}.

\item Roman V. Kiseliov made several suggestions for improving
the appearance of the book, and helped me with some \LaTeX~issues.

\item Alejandro C\'{e}spedes is working on the Spanish translation of this
book and found some typos.

\item Erich Nahum found a problem in my adaptation of Kenneth Reek's
  solution to the Sushi Bar Problem.

\item Martin Storsj\"{o} sent a correction to the generalized smokers problem.

\item Cris Hawkins pointed out an unused variable.

\item Adolfo Di Mare found the missing ``and''.

\item Simon Ellis found a typo.

\item Benjamin Nash found a typo, an error in one solution, and
a malfeature in another.

\item Alejandro Pulver found a problem with the Barbershop solution.

\end{itemize}
